---
title: "ANTsR Pipeline"
author: "Elizabeth Sweeney"
date: "August 21, 2014"
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{DICOM to NIfTI Conversion}
output:
  knitr:::html_vignette:
    toc: yes
---


```{r knit-setup, echo=FALSE, results='hide', eval=TRUE}
rm(list=ls())
library(knitr)
opts_chunk$set(cache = FALSE)
```


Install the ANTsR and ENARSC2015 R packages off of github using the library devtools.  Note that the ANTsR package may take a substantial amount of time to download. 


```{r, eval = FALSE, echo = TRUE}
library(devtools)
install_github('stnava/ANTsR')
install_github('muschellij2/ENARSC2015')
```

Load the ANTsR and ENARSC2015 libraries. 

```{r}
library(ANTsR)
library(ENARSC2015)
```

Load volumes
---

The output directory for the analysis (you should change this for your machine).

```{r set_enar_dir}
rootdir = "~/ENARSC2015"
rootdir = Sys.readlink(rootdir)
outdir <- file.path(rootdir, 'inst', 'ANTsR')
stopifnot(file.exists(outdir))
```

For this analysis, we provide the BRAINX data from OsiriX (http://www.osirix-viewer.com/datasets/) in the ENARSC2015 package.  The BRAINX data consists of multi-modal MRI from a patient with brain cancer.  Here we use the T1-w, T2-w and FLAIR volumes. 

Get the path for the T1-w, T2-w, and FLAIR volumes that are contained in the ENARSC2015 package. 

```{r get_img_names}
t1.path <- system.file("BRAINIX/NIfTI/T1.nii.gz", package = 'ENARSC2015')
t2.path <- system.file("BRAINIX/NIfTI/T2.nii.gz", package = 'ENARSC2015')
flair.path <- system.file("BRAINIX/NIfTI/FLAIR.nii.gz", package = 'ENARSC2015')
```

Load the T1-w, T2-w, FLAIR as an antsImage.

```{r read_imgs}
t1 <- antsImageRead(t1.path, 3)
t2 <- antsImageRead(t2.path, 3)
flair <- antsImageRead(flair.path, 3)
```

Plot volumes
---

View the original T1, T2 and FLAIR volumes.

```{r plot_imgs, fig.width= 10, fig.height = 3}
par(mfrow = c(1,3))
image(as.array(t1)[,,10], col = gray(0:64/64))
image(as.array(t2)[,,10], col = gray(0:64/64))
image(as.array(flair)[,,10], col = gray(0:64/64))
```

N3 Correction
---

Create volumes for the N3 corrected T1, T2 and FLAIR volumes.

```{r clone_imgs}
t1N3 <- antsImageClone(t1)
t2N3 <- antsImageClone(t2)
flairN3 <- antsImageClone(flair)
```

Perform the N3 bias feild correction on the T1, T2 and FLAIR volumes.

```{r n3_correct}
t1N3 = n3BiasFieldCorrection(t1, "4")
t2N3 = n3BiasFieldCorrection(t2, "4")
flairN3 = n3BiasFieldCorrection(flair, "4")
```

Write the N3 corrected files. 

```{r antsWrite}
antsImageWrite(t1N3, file.path(outdir, 'T1_ANTsR_N3Correct.nii.gz'))
antsImageWrite(t2N3, file.path(outdir,'T2_ANTsR_N3Correct.nii.gz'))
antsImageWrite(flairN3, file.path(outdir,'FLAIR_ANTsR_N3Correct.nii.gz'))
```

View the N3 bias feild correction on the T1, T2 and FLAIR volumes.

```{r plot_n3_imgs, fig.width= 10, fig.height = 3}
par(mfrow = c(1,3))
image(as.array(t1N3)[,,10], col = gray(0:64/64))
image(as.array(t2N3)[,,10], col = gray(0:64/64))
image(as.array(flairN3)[,,10], col = gray(0:64/64))
```

Templates
---

Read the template with and without skull. 

```{r read_template}
template.path <- system.file("Template/MNI152_T1_1mm_brain.nii.gz", package = 'ENARSC2015')
template.skull.path <- system.file("Template/MNI152_T1_1mm.nii.gz", package = 'ENARSC2015')
template <- antsImageRead(template.path, 3)
template.skull <- antsImageRead(template.skull.path, 3)
```

Plot the templates. 

```{r plot_template,  fig.width= 6.5, fig.height = 3}
par(mfrow = c(1,2))
image(as.array(template)[,,90], col = gray(0:64/64))
image(as.array(template.skull)[,,90], col = gray(0:64/64))
```


Registration  
---

First we will register the T1-w volume to the FLAIR volume. 

```{r register_rigid}
antsRegOut <- antsRegistration(fixed = flairN3, moving = t1N3 , typeofTransform = "Rigid",  outprefix = "./test")
t1.to.flair <-antsImageClone(antsRegOut$warpedmovout)
```


Write the registered T1-w volume. 

```{r write_rigid}
antsImageWrite(t1.to.flair, file.path(outdir, 'T1_ANTsR_N3Correct_regFLAIR.nii.gz'))
```

Next we plot the registered FLAIR volume and the T1-w volume to visually assess the quality of the registration. 

```{r plot_rigid, fig.width= 6.5, fig.height = 10}
par(mfrow = c(3,2))
slices <- c(5, 10, 15)
for(i in slices){
image(as.array(flairN3)[,,i], col = gray(0:64/64))
image(as.array(t1.to.flair)[,,i], col = gray(0:64/64))
}
```

Subtract the FLAIR and the T1-w images and plot the result.  

```{r subtract_image}
subtraction.image <- antsImageClone(flairN3)
ImageMath( 3 , subtraction.image  , "-", flairN3 , t1.to.flair )
image(as.array(subtraction.image)[,,10], col = gray(0:64/128))
```

Read in a region of interest (ROI).  In this case we have a brain mask of a manual segmentation of a brain tumor. 

```{r roi_img}
roi.path <- system.file("BRAINIX/NIfTI/ROI.nii.gz", package = 'ENARSC2015')
roi <- antsImageRead(roi.path, 3)
```

Display the FLAIR and the ROI  

```{r plot_roi}
par(mfrow = c(1,2))
image(as.array(flairN3)[,,15], col = gray(0:64/64))
image(as.array(roi)[,,15], col = gray(0:64/64))
```

Perform a non-linear registration of the T1-w image to the MNI template with the skull.  Note that we remove the InverseWarp and Warp files generates by ANTsR, as these files are quite large. 

```{r register_syn}
outprefix = file.path(rootdir, "inst", 'ANTsR', "ants")
antsRegOut.nonlin <- antsRegistration(fixed = template.skull, moving = t1.to.flair, typeofTransform = "SyN",  outprefix = outprefix)
```

```{r register_syn_out}
print(antsRegOut.nonlin)
```

Extract the non-linear registration of the T1-w image to the MNI template with the skull.

```{r clone_t1}
t1.to.flair.to.template <-antsImageClone(antsRegOut.nonlin$warpedmovout)
```

Apply the registration transformation to the FLAIR and the ROI image. 

```{r warp_flair} 
flair.to.template <- antsApplyTransforms(fixed=template.skull , moving=flairN3 , transformlist=antsRegOut.nonlin$fwdtransforms , interpolator="Linear")
roi.to.template <-antsApplyTransforms(fixed=template.skull , moving=roi , transformlist=antsRegOut.nonlin$fwdtransforms , interpolator="Linear")
```

Threshold the interpolated ROI image so that it is binary. 

```{r warp_roi}
roi.to.template.threshold <-antsImageClone(roi.to.template)
ThresholdImage(3, roi.to.template, roi.to.template.threshold, .5, 1)
```

Plot the transformed FLAIR, thresheld ROI and template with skull. 

```{r plot_warped_imgs, fig.width= 10, fig.height = 3}
par(mfrow = c(1,3))
image(as.array(flair.to.template)[,,110], col = gray(0:64/64))
image(as.array(roi.to.template.threshold)[,,110], col = gray(0:64/64))
image(as.array(template.skull)[,,110], col = gray(0:64/64))
```

Write the transformed FLAIR, ROI and  thresheld ROI.

```{r write_warped_imgs}
antsImageWrite(flair.to.template, file.path(outdir, 'FLAIR_ANTsR_N3Correct_regMNI_SyN.nii.gz'))
antsImageWrite(roi.to.template , file.path(outdir, 'ROI_regMNI_SyN.nii.gz'))
antsImageWrite(roi.to.template.threshold, file.path(outdir, 'ROI_threshold_regMNI_SyN.nii.gz'))
```

```{r remove_warp_img}
file.remove(paste0(outprefix, "1InverseWarp.nii.gz" ) )
file.remove(paste0(outprefix, "1Warp.nii.gz" ) )
```
