---
title: "DICOM to NIfTI"
author: "John Muschelli"
date: "August 21, 2014"
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{DICOM to NIfTI Conversion}
output:
  knitr:::html_vignette:
    toc: yes  
---

# DICOM to NIfTI Conversion in R

```{r knit-setup, echo=FALSE, results='hide', eval=TRUE}
library(knitr)
opts_chunk$set(echo=TRUE, prompt=FALSE, message=TRUE, warning=TRUE, comment="", cache=TRUE)
# cache.path <- path.expand("~/Dropbox/Packages/ENAR_SC_2015/vignettes/DCM2NII_cache/")
# opts_chunk$set(cache.path=cache.path)
```

## Introduction

This vignette describes how to convert from DICOM format to NIfTI using the `oro.dicom` package and useful `R` functions to read and write NIfTI files using the `oro.nifti` package.

Let's first install these packages.
```{r, eval=FALSE}
install_github('muschellij2/ENARSC2015')
# install.packages(c("oro.dicom", "oro.nifti"))
```

## Reading in DICOM data

The workhorse function for reading in DICOM data is the `readDICOMFile` function in `oro.dicom`, where individual filenames are passed to be read.  The `readDICOM` function wraps this function and converts all DICOM files in a path into one list of data.  The `dicom2nifti` function takes the list of header information and array of values and creates a 3D `nifti` object.

```{r}
dcm2nii <- function(path, ## path to DICOM files
                    outfile = NULL, ## output filename
                    retimg = TRUE, # write file out to disk?
                    ... # additional options to \code{\link{readDICOM}}
                    ){
  dicom <- readDICOM(path, flipud = FALSE, ...)
  nifti <- dicom2nifti(dicom)
  write = !is.null(outfile)
  if (retimg){
    if (is.null(outfile)) {
      outfile = tempfile()
      outfile = nii.stub(outfile)      
    }
  } else {
    stopifnot(!is.null(outfile))
  }
  
  if (write){
    writeNIfTI(nifti, filename = outfile, 
               verbose = TRUE, gzipped = TRUE)
  } 
  return(nifti)
}
```

Let's specify an output directory: `~/Desktop/results` where the output images should be.

```{r readin,  warning=FALSE, cache=TRUE}
library(oro.dicom)
library(oro.nifti)

# output directory for nifti files
outdir = tempdir()
outdir = path.expand(outdir)

if (!file.exists(outdir)){
  dir.create(outdir, showWarnings =  FALSE)
}

# Data taken from http://www.osirix-viewer.com/datasets/
mods = c("T1", "T2", "FLAIR", "ROI")
hdrs = imgs = vector(mode="list", length= length(mods))
names(imgs) = names(hdrs) = mods
imod = 1
for (imod in seq_along(mods)){
  mod = mods[imod]
  dicom_path = system.file(file.path("BRAINIX/DICOM", mod), package="ENARSC2015")
  fname = file.path(outdir, mod)
  headers <- readDICOM(dicom_path, pixelData=FALSE)$hdr
  hdrs[[imod]] = dicomTable(headers)
  img = dcm2nii(dicom_path, retimg = TRUE, outfile = fname)
  imgs[[imod]] = img
}
```

Some warnings may occur for this data.  Some data may not be correctly converted if they were acquired in a non-standard way.

### Explanation of Code
In 

> `dicom_path = system.file(file.path("DICOM", mod), package="ENARSC2015")`

We grabbed the directory for the DICOM data from the `ENARSC2015` directory.  The directories are mapped such as `DICOM/T1`.  The `readDICOM` function takes this directory, parses the DICOM files, and returns a list with slots `hdr` and `img`:

> `headers <- readDICOM(dicom_path, pixelData=FALSE)$hdr`

and note that `readDICOM` will **recursively** go through folders by default.  We specified `pixelData=FALSE`, indicated that `img` should be empty and only the header (metadata) should be read in.  Changing `dicom_path` to the path of your DICOM data is required, as well as output filenames

The 

> `hdrs[[imod]] = dicomTable(headers)`

command used the `dicomTable` function to reshape the list of headers in to a `data.frame` with rows being images and columns being information form the header.

Let us look at the header:
```{r headers}
class(hdrs[["T1"]])
rn = rownames(hdrs[["T1"]])
head(rn, 2)
rownames(hdrs[["T1"]]) = basename(rn)
hdrs[["T1"]][1:5, 1:5]
```
The filenames are the rownames of the `data.frame`. 

If we look at the column names of the `data.frame`, we see that they are a combination of the key information from the DICOM and the field name.
```{r header2}
colnames(hdrs[["T1"]])[1:5]
```

You may want to strip these off and simply use the field names.
```{r cn}
head(gsub(".*-(.*)", "\\1", colnames(hdrs[["T1"]])), 10)
```
There may be cases that these are not unique and subsetting the columns you want is a better strategy.

The `dcm2nii` function specified above wraps the 2 steps of reading in the DICOM data into an object and then using `dicom2nifti` on the object to convert the multiple DICOM slices into one 3D volume.  The results is a `nifti` object that can be written using `writeNIfTI` from the `oro.nifti` package and `dcm2nii` returns this nifti object of the image. 

## nifti objects

The `imgs` list is a list of objects of class `nifti`.
```{r nif_class}
sapply(imgs, class)
imgs[["FLAIR"]]
```
The output shows some information about the `nifti` object, such as dimensions and pixel resolution.  These are S4 objects and must be referenced by the @ symbol for slots or functions that extract these slots.


```{r img_info}
head(slotNames(imgs[["FLAIR"]]))
imgs[["FLAIR"]]@dim
dim(imgs[["FLAIR"]])
imgs[["FLAIR"]]@pixdim
```

The array of values are located in the `.Data` slot, but most array operations can be used on a `nifti` object, such as taking values greater than zero and taking the mean to get the proportion of the image greater than 0 or simple subsetting and replacement as in arrays.

```{r img_op}
i = imgs[["FLAIR"]]
mean(i > 0)
i[i < 0 ]= NA
```
## Reading and Displaying in NIfTI data

Now that we have NIfTI files, we can read them in directly:

```{r ex_readin}
fname = file.path(outdir, "FLAIR")
img = readNIfTI(fname)
```

We can plot the data using the `orthographic` function from `oro.nifti`:
```{r ortho}
orthographic(img, text="FLAIR image")
image(img)
image(img, z = 10, plot.type="single")
```

If we have an array, we can create `nifti` object with the `nifti` command:
```{r makenii}
arr = img > 300
class(arr)
nim = nifti(arr, dim = dim(img), pixdim=pixdim(img))
orthographic(nim)
```

We can also present overlays on images.  Note we have to remove 0's from the image or it will color these as well.
```{r over, dev='png'}
library(scales)
nim[nim == 0] = NA
nim[nim == TRUE] = 1
orthographic(img, nim, col.y=alpha("red", 0.25))
```


